A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-06-20 tex/BiSig.tex
[ tex ] Add another point to elaborate on simple types

  In response to this difficulty, we develop a \emph{simple} yet \emph{general} and \emph{constructive} theory of bidirectional type synthesis where \emph{bidirectional type synthesis} refers to the mutual use of type checking and inference based on bidirectional typing rules to determine the type of a program with type annotations.
  First, our theory is simple yet general as it deals with any simple type theory and its bidirectional type system that can be specified by a signature, analogous to the role of grammar in parsing.
  Second, our theory is constructive because it is not only formalised in a proof assistant based on constructive type theory but also formulated positively in order to compute.
  Specifically, the `proof' of logical decidability of type synthesis is instantiated to a synthesiser for the simple type theory specified by a valid signature, analogous to a parser generator.
  We introduce the notion of bidirectionalisation to refine annotatability in bidirectional typing.
  Type synthesis and bidirectionalisation result in the trichotomy of raw terms, which models a practical type synthesiser that may throw an exception to locate missing annotations.
  By formalizing our general theory of bidirectional type synthesis constructively, we provide a correct-by-construction `type-checker generator' for simple type theories, taking the first step towards a more general theory of bidirectional type synthesis.
\end{abstract}

>>>>

  In response to this difficulty, we develop a \emph{simple} yet \emph{general} and \emph{constructive} theory of bidirectional type synthesis where \emph{bidirectional type synthesis} refers to the mutual use of type checking and inference based on bidirectional typing rules to determine the type of a program with type annotations.
  First, our theory is simple yet general as it deals with any simple type theory and its bidirectional type system that can be specified by a signature, analogous to the role of grammar in parsing.
  Second, our theory is constructive because it is not only formalised in a proof assistant based on constructive type theory but also formulated positively in order to compute.
  Specifically, the `proof' of logical decidability of type synthesis is instantiated to a synthesiser for the simple type theory specified by a valid signature, analogous to a parser generator.
  We introduce the notion of bidirectionalisation to refine annotatability in bidirectional typing.
  Type synthesis and bidirectionalisation result in the trichotomy of raw terms, which models a practical type synthesiser that may throw an exception and locate missing annotations.
  By formalizing our general theory of bidirectional type synthesis constructively, we provide a correct-by-construction `type-checker generator' for simple type theories, taking the first step towards a more general theory of bidirectional type synthesis.
\end{abstract}

### 2023-06-20 tex/sec1-intro.tex
[ tex ] Add another point to elaborate on simple types

Our assumptions and contributions:
\begin{enumerate}
  \item Assumption: Syntax-directed typing rules
  \item Practical: bidirectionalisation over annotatability, decidability over soundness and completeness.
  \item Generic: Type checking once and for all bidirectional type systems for simple types.
  \item Correct and Constructive: Fully formalised in Martin-Löf type theory (with Axiom K).
    LEM is not assumed, so a decidable statement has both computational and non-trivial logical readings.
\end{enumerate}

>>>>

Our restrictions:
\begin{enumerate}
  \item Assumptions: Why only simple types? Various ways to deal with polymorphism; consistent on simple types.
  \item Assumption: Syntax-directed typing rules
\end{enumerate}

Our contributions:
\begin{enumerate}
  \item Generic: Type checking for simple type theories;
  \item Computational formulation: Bidirectionalisation over annotatability; Decidability over soundness and completeness;
  \item Verified and Constructive: Fully formalised in Martin-Löf type theory (with Axiom K).
    LEM is not assumed, so a decidable statement has both computational and non-trivial logical readings.
\end{enumerate}
